Modal Verification Patterns for Systems

### 2025-06-16 paper.tex
+

\paragraph{Virtual Memory Managers}
\label{sec:backgroundonvmm}
\todo{change the text}
Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

@@

The key pieces of functionality a VMM must implement itself, or have available in another OS component (depending on design), are
adding a new page mapping (whether the mapped page contains zeros, file data, or swap data), and removing an existing page mapping.
While this initially sounds like relatively modest functionality whose implementation may be complicated by hardware subtleties, a number of more serious complexities must be handled.

For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly access memory given its physical address, for example when interacting with hardware devices. For this reason, most kernels contain an \emph{identity map} of all physical memory in a machine. In some cases this is a literal identity map, where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.

Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap. One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly phased out in the wake of Spectre
%\mytodo{Colin finish}

These dove-tail with functionality in the OS scheduler, which since it is responsible for dealing with multiple address spaces, must keep track of which virtual addresses are valid (and in what way) in which address spaces, where some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode process), while others are valid in all address spaces (e.g., kernel data structure pointers). The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

@@

>>>>

\paragraph{Virtual Memory Managers with Multiple Address-Spaces}
\label{sec:backgroundonvmm}
\todo{change the text}
Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

@@

The key pieces of functionality a VMM must implement itself, or have available in another OS component (depending on design), are
adding a new page mapping (whether the mapped page contains zeros, file data, or swap data), and removing an existing page mapping.
While this initially sounds like relatively modest functionality whose implementation may be complicated by hardware subtleties, a number of more serious complexities must be handled.
%
%For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly access memory given its physical address, for example when interacting with hardware devices. For this reason, most kernels contain an \emph{identity map} of all physical memory in a machine. In some cases this is a literal identity map, where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.
%
%Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap. One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly phased out in the wake of Spectre
%\mytodo{Colin finish}

The OS scheduler, since it is responsible for dealing with multiple address spaces, must keep track of which virtual addresses are valid (and in what way) in which address spaces, where some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode process), while others are valid in all address spaces (e.g., kernel data structure pointers). The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

@@

\subsection{Speculative Observation in Systems Software}
\label{sec:speculativeobservation}
